Comparing binsegRcpp with other implementations of binary segmentation for changepoint detection

### 2025-05-06 article.Rnw
figure heap

\begin{table*}
  \centering
  \begin{tabular}{ccccccc}
    Name & Loss & Distrib. & Center & Scale & Params & Complexity \\
    \hline
    \texttt{mean\_norm} & Square & Normal & yes & no & 1 & $O(S)$ \\
    \texttt{meanvar\_norm} & Neg. Log. Lik. & Normal & yes & yes & 2 & $O(S)$ \\
    \texttt{poisson} & Neg. Log. Lik. & Poisson & yes & yes & 1 & $O(S)$ \\
    \texttt{l1} & Absolute & Laplace & yes & no & 1 & $O(S\log S)$ \\
    \texttt{laplace} & Absolute & Laplace & yes & yes & 2 & $O(S\log S)$ \\
  \end{tabular}
  \caption{\label{tab:losses}
    TODO
  }
\end{table*}

@@

%data_vec <- c(5, 0, 20, 15, -10, 0)
For example, we consider a simple example with six data points.
<<>>=
ex6_df <- data.frame(value = c(1, -7, 8, 10, 2, 4))
ex6_binsegRcpp <- binsegRcpp::binseg("mean_norm", ex6_df$value, max.segments = 4)
ex6_binsegRcpp$splits[, .(segments, end, before.mean, after.mean,
  invalidates.index, invalidates.after)]
@
The output above is a table with four rows, one for each model size.
For a max number of segments $M$, this representation of model parameters uses $M$ rows, so is asymptotically linear, $O(M)$.
The first row represents the trivial model: one segment across all data points, with \code{end=6} equal to the index of the last data point.
The model parameter of the trivial model is \code{before.mean=3}, which is the mean of all of the data.
The other rows represent the segments that result from the recursive splitting algorithm.
For example, in the second row, \code{end=2} means that the first change-point results in a split between data points 2 and 3.
The \code{before.mean = -3} and \code{after.mean=6} values are the means of the resulting segments, before and after that split.
The other columns, \code{invalidates.index} and \code{invalidates.after}, indicate the previous means which are no longer valid, after the given split.
For example, in the second row, \code{invalidates.index=1} identifies the first row, and \code{invalidates.after=0} identifies the mean before the split (not after), which implies that \code{before.mean=3} is no longer valid, after performing the split in the second row.
In the third row, we see a new split at \code{end=4}, resulting in new parameters \code{before.mean=9} and \code{after.mean=3}, which invalidate the previous mean in row 2 after the split (\code{after.mean=6}).

>>>>

\begin{table*}
  \centering
  \begin{tabular}{ccccccc}
    Name & Loss & Distrib. & Center & Scale & Params & Complexity \\
    \hline
    \texttt{mean\_norm} & Square & Normal & yes & no & 1 & $O(S)$ \\
    \texttt{meanvar\_norm} & Neg. Log Lik. & Normal & yes & yes & 2 & $O(S)$ \\
    \texttt{poisson} & Neg. Log Lik. & Poisson & yes & yes & 1 & $O(S)$ \\
    \texttt{l1} & Absolute & Laplace & yes & no & 1 & $O(S\log S)$ \\
    \texttt{laplace} & Neg. Log Lik & Laplace & yes & yes & 2 & $O(S\log S)$ \\
  \end{tabular}
  \caption{\label{tab:losses}
    TODO
  }
\end{table*}

@@

%data_vec <- c(5, 0, 20, 15, -10, 0)
For example, we consider a simple example with six data points.
<<>>=
ex6_df <- data.frame(value = c(1, -7, 8, 10, 2, 4))
ex6_binsegRcpp <- binsegRcpp::binseg(
  "mean_norm", ex6_df$value, max.segments = 4)
ex6_binsegRcpp$splits[, .(segments, end, before.mean, after.mean,
  invalidates.index, invalidates.after)]
@
The output above is a table with four rows, one for each model size.
For a max number of segments $M$, this representation of model parameters uses $M$ rows, so is asymptotically linear, $O(M)$.
The first row represents the trivial model: one segment across all data points, with \code{end=6} equal to the index of the last data point.
The model parameter of the trivial model is \code{before.mean=3}, which is the mean of all of the data.
The other rows represent the segments that result from the recursive splitting algorithm.
For example, in the second row, \code{end=2} means that the first change-point results in a split between data points 2 and 3.
The \code{before.mean = -3} and \code{after.mean=6} values are the means of the resulting segments, before and after that split.
The other columns, \code{invalidates.index} and \code{invalidates.after}, indicate the previous means which are no longer valid, after the given split.
For example, in the second row, \code{invalidates.index=1} identifies the first row, and \code{invalidates.after=0} identifies the mean before the split (not after), which implies that \code{before.mean=3} is no longer valid, after performing the split in the second row.
In the third row, we see a new split at \code{end=4}, resulting in new parameters \code{before.mean=9} and \code{after.mean=3}, which invalidate the previous mean in row 2 after the split (\code{after.mean=6}).
